---
title: "STAT443 Assignment 5"
author: "Qing Ruan 4043939"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,error=T) 
```

 
## Question 1: Slice sampler
Use R to code a slice sampler that samples values from the density $f(y)\propto \exp(-\sqrt{y})$ for $y>0$. Construct a plot that illustrates the density and add points corresponding to the values chosen numerically (i.e., using integrate() and compare these values with simulated ones.) 

```{r,fig.width=0.9,eval=FALSE}
curve(exp(-x^(0.5)),from=0,to=100,ylab="Density")

slice<-function(x0,n){
  xx<-x0
  u<-runif(1,0,exp(-x0^(0.5)))
  yy<-u
  
  for (i in 1:(n-1)){
  x0<-runif(1,0,(-log(u))^2)
  xx<-c(xx,x0)
  u<-runif(1,0,exp(-x0^(0.5)))
  yy<-c(yy,u)
  }
  return(data.frame(xx=xx,yy=yy))
}

dat<-slice(x0=1,100000)
points(dat$xx,dat$yy,pch=16,cex=0.6)

exp(-2.17^(0.5))

```




## Question 2: Your own Metropolis
Suppose that $y_1,...,y_k$ are independent samples from $k$ binomial distribution each indexed by $n_i$ and with the logit of the probabilities $\pi_i$ given by

$$
\ln\left(\frac{\pi_i}{1-\pi_i}\right)=\beta_0+\beta_1x_i
$$

Construct and test a Metropolis within Gibbs algorithm for sampling from the posterior distribution $[\beta_0,\beta_1|\{y_i\},\{x_i\}]$.Note that by ``Metropolis within Gibbs'' we mean that (some) samples are drawn from their full-conditional distribution using the Metropoli-Hasting algorithm. This approach is useful when the full conditional distribution are not in a form that corresponds to known distributions. For priors on the parameters use $[\beta_i]=N(\mu_i,\tau_i)$ (mean and precision, i=0,1) and appropriate values of $\mu_i$ and $\tau_i$.


                          
## Question 3: Ouch! Logistic pain.

### 1. Fit a simple logistic regression model using standard methods and report the estimates, standard error and 95\% confidence interval for the parameters.

The binomail modal for the probability of reacting $\pi_i$ is 

$$
\ln\left(\frac{\pi_i}{1-\pi_i}\right)=\beta_0+\beta_1x_i=6.469-5.567x_i
$$
                          
standard error for $\beta_0$ and $\beta_1$ are 2.418, 2.044 respectively. 

The 95\% confident intervel for $\beta_0$ and $\beta_1$ are $[1.728560,11.208790]$ and $[-9.572126 ,-1.561398]$.


### 2. Report an estimate (and 95\% confidence interval) for the anesthetic concentration at which 90\% of patients show no reaction.

Given $\pi^*=1-0.9=0.1$, the estimate of the anesthetic concentration will be 

$$
\ln\left(\frac{0.1}{0.9}\right)=\hat{\beta_0}+\hat{\beta_1}\hat{x}=6.469-5.567\hat{x}\\
\Rightarrow \hat{x}= \frac{\left(\ln\left(\frac{0.1}{0.9}\right)-\hat{\beta_0} \right)}{\hat{\beta_1}}=1.556714
$$

By the properties of MLEs, we know asymptotically, $\mathbf{\beta}=\begin{pmatrix}\hat{\beta_0}\\\hat{\beta_1}\end{pmatrix}$ is normallly distributed with variance-covariance matrix 

$$
V(\hat{\beta})=\begin{pmatrix}
5.848999 &-4.848174\\
-4.848174 & 4.176262
\end{pmatrix}
$$

Using the Multivariate Delta method, we have 
$$
\frac{\hat{x}-g(\hat{\beta})}{Var(g(\hat{\beta}))}\sim N(0,1))
$$

where $Var(g(\hat{\beta}))=g'V(\hat{\beta})g$ 

$$
\hat{x}'=g'(\beta)=\begin{pmatrix}
\frac{\partial g(\beta)}{\partial \beta_0}\\
\frac{\partial g(\beta)}{\partial \beta_1}
\end{pmatrix}
=\begin{pmatrix}
-\frac{1}{\beta_1}\\
-\frac{\ln(0.1/0.9)-\beta_0}{\beta_1^2}
\end{pmatrix}
=\begin{pmatrix}
-\frac{1}{\hat{\beta}_1}\\
\frac{\hat{\beta}_0-\ln(0.1/0.9)}{\hat{\beta}_1^2}
\end{pmatrix}
=\begin{pmatrix}
0.1796377\\ 0.2796458
\end{pmatrix}
$$

$$Var(g(\hat{\beta}))=g'V(\hat{\beta})
=\begin{pmatrix}
0.1796377\\ 0.2796458
\end{pmatrix}'\begin{pmatrix}
5.848999 &-4.848174\\
-4.848174 & 4.176262
\end{pmatrix}
=\begin{pmatrix}
0.1796377\\ 0.2796458
\end{pmatrix}=0.02824131
$$ 

Therefore the 95\% confidence interval of $\hat{x}=1.556714$ is $[1.227347,1.886097]$.  



```{r,warning=FALSE,fig.height=3,fig.width=5}
# This dataset contains the following columns:

# move is a binary numeric vector coded for whether the patient reacted 
# (0 = no reacting, 1 = reacting)

# conc is anesthetic concentration
library(DAAG)
library(knitr)
(z <- table(anesthetic$move, anesthetic$conc))

tot <- as.numeric(apply(z,2,sum))
prop <- z[2,]/tot
conc <- as.numeric(dimnames(z)[[2]])

# plot
plot(conc, prop, xlab = "Concentration", ylab = "Proportion",
     xlim = c(.5,2.5), ylim = c(0, 1), pch = 16)
chw <- par()$cxy[1]
text(conc - 0.75 * chw, prop, paste(tot), adj = 1)
abline(h = oprop, lty = 2)

# GLM logistic model
anes.logit <- glm(cbind(move,nomove) ~ conc, family = binomial(link = logit),
data = anesthetic)

summary(anes.logit)

# CI for beta_0 & beta_1
# same as anes.logit$coeff+c(-1,1)*qnorm(0.975)*se
# confint() gives profile confidence interval instead
confint(anes.logit)
confint.default(anes.logit)       

(beta<-as.numeric(anes.logit$coeff))


# MLE estimate of conc.new
(conc.new<-(log(0.1/0.9)-beta[1])/beta[2])

# var-cov matrix
(V=vcov(anes.logit))

(g <- c(-1/beta[2],(beta[1]-log(0.1/0.9))/(beta[2]^2)))
(tau2 <- t(g)%*%V%*%g)

# MLE CI for concentration
conc.new+c(-1,1)*qnorm(0.975)*sqrt(tau2)
```

### 3. Either using BUGS/JAGS or computer code based on the algorithm outlined in Question 2 repeat the analysis carried out in part 1) and 2) using vague prior distirubtion for the parameter. 

Let $y_i$ be the number of patients reacting, $x_i$ be anesthetic concentration, the model we use 

$$y_i\sim\text{Bin}(\text{tot}_i,\pi_i)$$
$$\text{logit}(\pi_i)=\beta_0+\beta_1\text{conc}_i \sim N(0,1e+12)$$


```{r,warning=FALSE,echo=FALSE,message=FALSE}
library(R2jags)

# model
anesthetic.model = function(){
  # data model 
    for (i in 1:6){
        y[i] ~ dbin(p[i],n[i])
        logit(p[i]) <- beta[1] + beta[2]*conc[i]
    }
  # prior 
    for (i in 1:2){beta[i]~dnorm(0,1e-6)}
  # predict
    conc.new <- (logit(0.1)-beta[1])/beta[2]
}
  

# initial values 
inits=function(){
  list('beta'=rnorm(2,0,1))}

params=c('beta','conc.new')

y = as.numeric(z[2,])
n = tot
conc = conc 

data = c('y','n','conc')

# output
jagsfit1=jags(data=data,inits=inits,params,n.iter=100000,
              model.file=anesthetic.model)
print(jagsfit1)

jags.mcmc1=as.mcmc(jagsfit1)

# diagnostic plot
par(mfrow=c(2,2))
traceplot(jags.mcmc1,col=topo.colors(3),lty=1,lwd=2)
gelman.diag(jags.mcmc1)


coda1=jagsfit1$BUGSoutput$sims.matrix
plot(density(coda1[,3],from=0,to=5),main = "Distribution of conc.new")
```

The estimate of the anesthetic concentration at wich 90\% of patients show no action is 1.52. And the credible interval is given by $[1.31,2.09]$.


### 4. Commment on likely reasons for differences betwwen Bayesian and MLE estimates.

* MLEs and Bayesian had similar estimates for **conc.new** since we assume a vague prior for the Bayesian model. (MLE gave the mean. Bayesian gave both mean and median but the distribution of **conc.new** is skewed so we tend to use the median value.). 

* The MLE confidence interval estimators are developed based on the asymptotic property of **conc.new** being normally distributed, while Bayesian estimation treats **conc.new** itself as a random variable that does not rely on the same assumpation. The density function shows  **conc.new** is not normally distributed. Therefore, Bayesian and MLE resulted in different estimates for the 95\% interval for **conc.new**.


## Question 4: Meta-analysis

In a meta-analysis, results from a number of studies are summarised using a statistical model.
The datset 'Beta blockers' provides data from 22 clinical trials that were carried out to assess the effectiveness of beta-blockers administered to heart attack patients.

Use JAGS to carry out an analysis of these data that summarises the effect of the beta-blocker on the survival prospects of heart attack patients. Write up your analysis in a way that could be understood by the researchers carrying out the 22 studies.

Let $y_{0j}$ and $y_{1j}$ be the number of deaths in study $j$ and $p_{0j}$ and $p_{1j}$ denote the probabilities of death in control (**group=0**) and treatment group (**group=1**). We interest in the logarithm of the odds ratio, $\text{log(OR}_{j})=\log\left(\frac{p_{1j}/1-p_{1j}}{p_{0j}/1-p_{0j}}\right)$. One approach to estimate odds ratio and the corresponding sampling variance provided by Gelman et al (2004, p147, Eq.(5.23) and Eq.(5.24)). The estimated log-odds ratios and standard errors are listed in the table below.

$$
OR_{j}=\log\left(\frac{y_{1j}}{n_{1j}-y_{1j}}\right)-\log\left(\frac{y_{0j}}{n_{0j}-y_{0j}}\right)
$$

$$
\sigma^2_j=\frac{1}{y_{1j}}+\frac{1}{n_{1j}-y_{1j}}+\frac{1}{y_{0j}}+\frac{1}{n_{0j}-y_{0j}}
$$

We regard the studies are exchangeable and assume the log(OR)’s are normally distributed. The model is 

1. $\text{OR}_{j}|\theta_j\sim N(\theta_{j},\sigma_{j}^{2})$

2. $\theta_j\sim N(\mu,v^{2})$

3. $\mu\sim N(0,1e+12)$ and $v^{2} \sim U(0,10)$


```{r}
beta_blockers<-read.table("BetaBlockers2.txt", header=T, sep="\t")

beta_blockers$group<-factor(beta_blockers$group)

beta_blockers$Study<-factor(beta_blockers$Study)

library(dplyr)
library(xtable)

dat0 <- beta_blockers %>% filter(group==0)%>%
  select(Study,deaths0=deaths,total0=total) %>%
  mutate(survive0=total0-deaths0,odd0=deaths0/survive0)
dat1 <- beta_blockers %>% filter(group==1)%>%
  select(Study,deaths1=deaths,total1=total) %>%
  mutate(survive1=total1-deaths1,odd1=deaths1/survive1)

bb<- dat0%>%inner_join(dat1)%>%mutate(log_OR=log(odd1/odd0),
                           SE=sqrt(1/deaths1+1/survive1+1/deaths0+1/survive0))


kable(bb)

# model
b1.model = function(){
  # data model 
   for (j in 1:22) {
     P[j] <- 1/sig2[j]  # precision of each treatment effect
     Y[j] ~ dnorm(theta[j],P[j])
     theta[j] ~ dnorm(mu,tau)} 
  # 1/tau describe the variability in log odds ratios between studies
  # prior 
    mu ~ dnorm(0,1e-06) # the mean logOR of jth study
    tau <- 1/v2        # the precision of jth study
    v2 ~ dunif(0,10)     
  # estimate
    OR <- exp(mu) 
}
  
bdata <- list(Y=as.numeric(bb$log_OR),sig2=(as.numeric(bb$SE))^2)
bpar <- c("theta","v2","tau","OR")
binits<-function(){list("mu"=c(0),"v2"=runif(1))}

set.seed(100)

jagsfit2<-jags(data=bdata, inits=binits, parameters.to.save=bpar, 
     n.iter=10000,model.file=b1.model)

jagsfit.mcmc2=as.mcmc(jagsfit2)

coda=jagsfit2$BUGSoutput$sims.matrix

(stats=summary(jagsfit.mcmc2))

# 95\% credible interval for each study

plot(1:22,seq(-0.8,0.6,length.out = 22),type="n", xlab="Studies",ylab="OR",bty="n",axes=F)
axis(side = 1, tck=-0.015, line=-0.4,at = c(seq(from=1,to=22,by=1)))
axis(side = 2, tck=-0.015, line=-0.4,at = c(seq(from=-0.8,to=0.6,by=0.2)))

  for(i in 1:22){ 
    points(i,stats$stat[i+5,1],pch=19)
    points(i,Y[i],pch=1)
    lines(c(i,i),c(stats$quan[i+5,1],stats$quan[i+5,5]),lwd=1)
  }

abline(0,0,lwd=0.2)
legend(1.1,0.6,c("Posterior mean","MLE"),pch=c(19,1),bty="n")

# another way to model 
b2.model=function(){
       for( i in 1:22) {
        r0[i] ~ dbin(p0[i], n0[i])
        r1[i] ~ dbin(p1[i], n1[i])
        logit(p0[i]) <- mu[i]
        logit(p1[i]) <- mu[i] + delta[i] + study[i]*beta
        theta[i] <- logit(p1[i])-logit(p0[i])
        mu[i] ~ dnorm(0,1e-6)
        delta[i] ~ dnorm(d, prec)
       }
       OR <- exp(d)
       d ~ dnorm(0.0,1.0E-6)
      tau~dunif(0,10)
      tau.sq<-tau*tau
      prec<-1/(tau.sq)
      beta ~ dnorm(0.0,1.0E-6) #additional prior for beta 
    }

r0<-bb$deaths0
r1<-bb$deaths1
n0<-bb$total0
n1<-bb$total1
study<-bb$Study

bdata <- list(r0=r0,n0=n0,r1=r1,n1=n1,study=study)
bpar <- c("d","tau","OR","theta")

binits<-function(){list("d"=runif(1),"tau"=c(1))}

jagsfit22<-jags(data=bdata, inits=binits, parameters.to.save=bpar, 
     n.iter=10000,model.file=b2.model)

jagsfit22

stats=summary(as.mcmc(jagsfit22))

# 95\% credible interval for each study

plot(1:22,seq(-0.8,0.6,length.out = 22),type="n", xlab="Studies",ylab="log Odds Ratio",bty="n",axes=F)
axis(side = 1, tck=-0.015, line=-0.4,at = c(seq(from=1,to=22,by=1)))
axis(side = 2, tck=-0.015, line=-0.4,at = c(seq(from=-0.8,to=0.6,by=0.2)))

  for(i in 1:22){ 
    points(i,stats$stat[i+5,1],pch=19)
    points(i,Y[i],pch=1)
    lines(c(i,i),c(stats$quan[i+5,1],stats$quan[i+5,5]),lwd=1)
  }

 abline(stats$statis[2,1],0,lwd=0.2)
legend(1.1,0.6,c("Posterior mean","MLE"),pch=c(19,1),bty="n")

```




## Question 5: Trout data
For the West Coast trout data set (file wctrout_mis.txt), some of the covariates are missing. Use JAGS to fit the model:

$$
\text{logit}(p_i)=\beta_0+\beta_1S_i+\beta_2L_i+\beta_3S_iL_i
$$

where $S_i$ is the indicator for the sex of the fish (0=male,1=female) and $L_i$ is its standardised length. Also predict the missing sex for fish 1 and the missing length for fish 2. Note that the lengths have been standardised already and that there are 1961 fish in the file. **Hint** you will need to model both length and sex, something not normally required for predictors in a regression analysis.


```{r,eval=FALSE}
head(tick)
mean(tick)

# model
tickmodel<-function(){
  # data model 
  for (i in 1:length(tick)){
    tick[i]~dpois(lambda)
  }
  # prior 
  lambda~dgamma(0.001,0.001)
}

# initial values 
inits=function(){
  list('lambda'=rgamma(1,1,1))
}

params=c('lambda')

# output
jagsfit1=jags(data=list(tick=tick),inits=inits,
              params,n.iter=100000,model.file=tickmodel)
(summ=jagsfit1$BUGSoutput$summary)


jags.mcmc1=as.mcmc(jagsfit1)

# diagnostic plot
par(mfrow=c(1,2))
traceplot(jags.mcmc1,col=topo.colors(3),lty=1,lwd=2)
gelman.diag(jags.mcmc1)
```

The traceplot and the BGR diagnostic indicates that the MCMC converges (iteration=5000). With a gamma prior with $\alpha_0=0.001$ and $\beta_1=0.001$, a 95% credible interval from the posterior distribution for $\lambda$ is in the range (approximate) [6.01,7.12] with expected value about 6.56.


## Question 6: Posterior model probabilities
Link and Barker (2010) give an example where two models are considered for the following binomial data: $y=(8,16)'$ and $n=(20,30)'$. The two models are:
 add changes 

